A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-03-02 analysis/paper/paper.Rmd

`r anchor("fig", "results_overview")`
![](`r figure_file("overview.svg", experiment_id="9-main_figure")`)
`r ref("fig", "results_overview")` HERE ARE THE RESULTS

Overall, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree CREF. These results are consistent between real and synthetic datasets CREF, confirming both the relevance of the synthetic data and the accuracy of the real datasets' gold and silver standards. However, results are more variable when comparing the three different evaluation metrics, 
##

The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset. In some 
## Prior information

@@

@@

`r anchor("fig", "implementation_overview")`
![](`r figure_file("method_components.svg", experiment_id="4-method_characterisation")`)
`r ref("fig", "implementation_overview", anchor=TRUE)` First characterization of each method.

@@

```{r}
add_sfig(
figure_file("method_components.svg","4-method_characterisation"), 
"method_components", 
"TI methods frequently share steps"
)
```
## Evaluating the accuracy of the inferred trajectory

When looking across all datasets, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics. Over time, new methods

Top methods tend to contain similar components, with …

Over time, both the average and maximum performance increased with the addition of new methods `r ref("sfig", "top_performers_over_time")`

Performance was stable between real and synthetic datasets `r ref("sfig", "real_synthetic_comparison")`

```{r}
add_sfig(
        figure_file('real_synthetic_comparison.svg', "5-optimise_parameters/5-real_synthetic_comparison")
, 
"real_synthetic_comparison", 
"Comparison of method performance between real and synthetic datasets. ", 
12, 
4
)
```
# Discussion
<!-- Some intro -->
Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list created by Sean Davis [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) . It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice. In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.

<!-- Results  -->
Based on the results of our benchmark, we created a set of practical guidelines for method users `r ref("fig", "user_guidelines")`. As method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data. For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure. In each case, a different set of methods performs optimally, with Monocle DDRTree performing best when the data contains a complex tree, while Slingshot performing better when the trajectory could as well be less complex. When the data is expected to contain multiple trajectories or cycles, no methods were included in the current version of the evaluation, although AGA `r cite_methods("aga")` for disconnected trajectories, and Waddington-OT `r cite_methods("wadot")` and AGA `r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories. If the user knows the exact topology, our evaluation suggests reCAT for cycles, SCORPIUS for linear trajectories, and slingshot for bifurcating trajectories, although in the last case other topologies could also be returned by the method if it better fits the data. When the users knows the exact . For cases where the user does

@@

```{r}
add_sfig({
figure_file("edge_flip_comparison.svg", experiment_id="3-score_rules/4-edge_flip")
}, 
"edge_flip_comparison", 
"Edge flip score values for common trajectory topologies", 
15, 
5
)
```

# Acknowledgements

@@

```{r}
git2r::repository(".")
```
I think we should mention something here about sc transcriptomics, flow/mass cytometry etc...
For example, I talked with Abhishek the other day and his view was really flow oriented. I think people who primarily use flow will be confused by whether this paper looks at scRNAseq or both flow and scRNAseq
Love this figure! It's a great summary of the /whole/ study methodology
I have several comments (that I might address myself if I get to it first). I'll post them as separate comments, so they can get resolved individually.
I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model
This is complex though to put into this figure :p. Have a look at 4-method_characteristics/wrapping.svg I think it is the least crucial of all parts of the figure now, so.. But feel free to try it yourself :p
a handful of the most recent methods
Repetition of most: put "The majority of" instead?
..

>>>>

Overall, we find that slingshot predicts the most accurate trajectories across different data sources, trajectory types and metrics, followed by TSCAN and Monocle DDRTree (`r ref("fig", "results_overview")`). These results are consistent between real and synthetic datasets (`r ref("fig", "results_overview")` and `r ref("sfig", "real_synthetic_comparison")`), confirming both the relevance of the synthetic data and the accuracy of the real datasets' gold and silver standards. However, results are more variable when comparing the three different evaluation metrics. Some methods, mainly those able to detect more complex trajectory types, have higher edge flip scores. Other methods, not necessarily those with the highest edge flip score, are better at preserving the ordering and neighbourhood of the cells.

`r anchor("fig", "results_overview")`
![](`r figure_file("overview.svg", experiment_id="9-main_figure")`)
`r ref("fig", "results_overview")` HERE ARE THE RESULTS
## Trajectory types

The main distinguishable feature among TI methods is the extent to which a method can adapt its trajectory topology to a dataset. Most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "implementation_overview", "a")`). However, only a handful, mostly recent, methods can handle more complex graph topologies which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).

Evaluation across trajectory types

## Prior information

@@

```{r}
add_sfig(
figure_file("method_components.svg","4-method_characterisation"), 
"method_components", 
"TI methods frequently share steps"
)
```

Top methods tend to contain similar components. Principal curves…

@@

@@

# Discussion
<!-- Some intro -->
Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. Indeed, for no other single-cell analysis types have so many tools been developed, according to omictools.org [@henry_omictoolsinformativedirectory_2014] (https://omictools.com/single-cell-rna-seq-category), the "awesome single cell software" list created by Sean Davis [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools [@zappia_exploringsinglecellrnaseq_2017] (https://www.scrna-tools.org) . It is therefore critical that these methods, now reaching `r sum(methods$is_ti, na.rm=TRUE)`, are evaluated to guide the users in their choice. In this preprint, we summarise the initial version of our evaluation of these methods, focusing on the quality control and the accuracy of their model using their default parameters on real and synthetic data.

<!-- Results  -->
Based on the results of our benchmark, we created a set of practical guidelines for method users `r ref("fig", "user_guidelines")`. As method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data. For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure. In each case, a different set of methods performs optimally, with Monocle DDRTree performing best when the data contains a complex tree, while Slingshot performing better when the trajectory could as well be less complex. When the data is expected to contain multiple trajectories or cycles, no methods were included in the current version of the evaluation, although AGA `r cite_methods("aga")` for disconnected trajectories, and Waddington-OT `r cite_methods("wadot")` and AGA `r cite_methods("aga")` for regular graphs are currently the only methods in literature able to handle these types of trajectories. If the user knows the exact topology, our evaluation suggests reCAT for cycles, SCORPIUS for linear trajectories, and slingshot for bifurcating trajectories, although in the last case other topologies could also be returned by the method if it better fits the data. The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory. Here, to our knowledge, only ElPiGraph.R (https://github.com/Albluca/ElPiGraph.R), which is not yet published, can be used.

@@

```{r}
add_sfig({
figure_file("edge_flip_comparison.svg", experiment_id="3-score_rules/4-edge_flip")
}, 
"edge_flip_comparison", 
"Edge flip score values for common trajectory topologies", 
15, 
5
)
```
# Acknowledgements

@@

```{r}
git2r::repository(".")
```
I think we should mention something here about sc transcriptomics, flow/mass cytometry etc...
For example, I talked with Abhishek the other day and his view was really flow oriented. I think people who primarily use flow will be confused by whether this paper looks at scRNAseq or both flow and scRNAseq
Love this figure! It's a great summary of the /whole/ study methodology
I have several comments (that I might address myself if I get to it first). I'll post them as separate comments, so they can get resolved individually.
I would include a part for the method wrappers; that is the way that linear, cellcluster, cellgraph and generic TI's get translated to the probabilistic trajectory model
This is complex though to put into this figure :p. Have a look at 4-method_characteristics/wrapping.svg I think it is the least crucial of all parts of the figure now, so.. But feel free to try it yourself :p
Should we create a separate github containing a readme, referencing to the google doc, for this?
..
TODO
TODO
to do: add metric figure
needs to be updated with waypoints
this is not used anymore
